Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-12-16 manuscript/prb-submitted/paper.tex
revised paper JT

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-bonds-labelled-2}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-cubic-bonds}
%\includegraphics[height=5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cell-pbnm-15-15-bonds}\\
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-cubic-final}
%\includegraphics[height=4.5cm,keepaspectratio]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/cages-rotated-15-15-final}
\caption{
Crystal structures and coordination cages.
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-beta-alpha-0-width-row}
\caption{
Splitting in $t_{2g}$ manifold induced by the
octahedral rotations.
%The changes in the energies of $t_{2g}$ states 
$E_{xy},E_{yz},E_{zx}$ 
%and the average of $t_{2g}$ energy
and $E_0^{t_{2g}}$
as a function of the rotation angel $\beta$
at $\alpha=0$ (no tilting),
$\Delta_o=4q_o/15a$ ($D_4=1/200$), $D_2=\sqrt{D_4}$,
and three values of $q_A/q_o$.
(a)
The B cage dominates at $q_A/q_o=1/2$;
it lowers $xy$ state and raises $yz,zx$ states
with increasing $\beta$.
As $q_A/q_o$ increases to $1$, 
(b)
A cage almost cancels the effect of B cage reducing the splitting,
while it dominates at $q_A/q_o=2$ (c) where it switches the order of the states. 
}
\label{fig:t2g-alpha0}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-alpha-beta-0-width-row-character}
\caption{
Splitting and mixing in $t_{2g}$ manifold induced by octahedral tilting.
(a-c) 
$E_D, E_{\pm}$ and $E_0^{t_{2g}}$
as a function of $\alpha$
at 
$\beta=0$;
other parameters are the same as in Fig.~\ref{fig:t2g-alpha0}.
(d-f) Weight $W_{xy}$ of $xy$ orbital in 
$\ket{E_{\pm}}$ states 
shown in (a-c).
The $xy$ orbital predominantly
contributes to 
$\ket{E_{+}}$
at $q_A/q_o=1/2,1$ (a,b,d,e), 
but to   
$\ket{E_{-}}$
at $q_A/q_o=2$ (c,f).
}
\label{fig:t2g-beta0}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-beta-alpha-10-width-row-beta-0-15}
\caption{
Splitting in $t_{2g}$ manifold due to octahedral rotation and tilting.
$t_{2g}$ energies as a function of $\beta$ at $\alpha=10^\circ$
at the same values of other parameters as in Figs.~\ref{fig:t2g-alpha0},\ref{fig:t2g-beta0}.
There are two anticrossings between the three states around $\beta\sim\alpha$.
The character of the states depends on 
the competition between the two angles and 
the two ion cages.
}
\label{fig:t2g-ge-Es}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/wts-energy-vs-beta-alpha-10-width-row-beta-0-15}
\caption{
The dependence of the character of the $t_{2g}$ states
on the competitions between
$\alpha,\beta$ and 
A,B ion cages.
Weights of the three $t_{2g}$ orbitals in the three 
$t_{2g}$ eigenstates 
shown in
Fig.~\ref{fig:t2g-ge-Es}.
%(at the same parameters).
At $q_A=q_o/2$, B cage dominates and
$\ket{E_1}$ is predominantly $yz,zx$ at $\beta\sim 0\ll\alpha=10^\circ$, but 
becomes $\sim xy$ at $\beta\gtrsim10^\circ$ (a).
The situation flips when A cage dominates at $q_A=2q_o$ (d).
Similarly, the character of $\ket{E_3}$ at 
$q_A=q_o/2$
resembles 
that of $\ket{E_1}$ at $q_A=2q_o$  [compare (c) with (d)], and vice versa [compare (f) with (a)].
The character of the middle state $\ket{E_2}$ (b,e)
becomes $\sim xy$ only around the ``resonance'' region
$\beta\sim \alpha$.
}
\label{fig:t2g-ge-wts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-wts-eg}
\caption{
Energies and characters of $e_g$ states.
$E_{\pm}$ and $E_0^{e_g}$
as a function of $\beta$ at $\alpha=10^\circ$, 
and the same parameters as in Fig.~\ref{fig:t2g-ge-Es}.
At $q_A=q_o/2$ (a,c) where B cage dominates,
$\alpha$ tends to make $\ket{E_{+}}$
more $x^2-y^2$ (see small $\beta$ region),
while $\beta$ tends to do the opposite 
(see large $\beta$ region).
The roles of $\alpha,\beta$ switch
at $q_A=2q_o$ (b,d), where A cage dominates.
We see that the splitting is relatively small as compared to $t_{2g}$ manifold.
Approximate results obtained by a second order expansion of the Hamiltonian given in Eqs.\ref{eq:eg-hiix}-\ref{eq:eg-E0}
are also shown for comparison.
}
\label{fig:e-wts-eg}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/full-lattice-potential-tests/full-lattice-vs-AB}
\caption{
The effect of the potential of full ionic lattice.
$t_{2g}$ spectrum computed using the potential of full lattice 
and that obtained from only nearest AB cages (and the anion octahedron itself)
are
shown as a function of $q_A/q_o$ at 
$\alpha=0,\beta=20^\circ$, $\Delta_o=4q_o/15a$.
Full lattice only shifts the balance between the two cation cages towards A 
(i.e, the crossing point, where the two cages balance each other, moves to smaller $q_A$)
but keeps the qualitative picture the same.
}
\label{fig:t2g-lattice}
\end{center}
\end{figure}

@@

Of course, beside the crystal field we consider in our simplistic model,
in real materials,
 there would be other local effects like
electronic correlations (Hubbard and Hund's interaction)
and spin-orbit interaction, 
as well as band structure effects arising from
the electron hopping between the B ions.
The interplay between these and the crystal field (or JT) effect 
we discussed in this manuscript is also an interesting future work.

>>>>

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/struc/structures-combined-bonds-labelled-2}
\includegraphics[width=1\linewidth]{structures-combined-bonds-labelled-2}
\caption{
Crystal structures and coordination cages.
Crystal structures of cubic (a) and orthorhombic (b) perovskites.
The anions (red) form the shown octahedra around 
the B cations (gold), while the A cations (green) sit in the voids between the octahedra.
In the orthorhombic structure shown here (b),
 the octahedra are rotated and tilted by $15^\circ$.
(c-d)
The coordination cages of A %(green) 
and B 
%(golden) 
cations just next to the anion 
%(red) 
octahedron
around a B ion in cubic %(c)
 and orthorhombic %(d)
  perovskites.
In the anion octahedron's frame,
the cages are symmetric and aligned to the octahedron in the cubic perovskites (c),
while they are rotated and slightly distorted in the orthorhombic perovskites (d).
}
\label{fig:struc}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-beta-alpha-0-width-row}
\includegraphics[width=1.0\linewidth]{energy-vs-beta-alpha-0-width-row}
\caption{
Splitting in $t_{2g}$ manifold induced by the
octahedral rotations.
%The changes in the energies of $t_{2g}$ states 
$E_{xy},E_{yz},E_{zx}$ 
%and the average of $t_{2g}$ energy
and $E_0^{t_{2g}}$
as a function of the rotation angel $\beta$
at $\alpha=0$ (no tilting),
$\Delta_o=4q_o/15a$ ($D_4=1/200$), $D_2=\sqrt{D_4}$,
and three values of $q_A/q_o$.
(a)
The B cage dominates at $q_A/q_o=1/2$;
it lowers $xy$ state and raises $yz,zx$ states
with increasing $\beta$.
As $q_A/q_o$ increases to $1$, 
(b)
A cage almost cancels the effect of B cage reducing the splitting,
while it dominates at $q_A/q_o=2$ (c) where it switches the order of the states. 
}
\label{fig:t2g-alpha0}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-alpha-beta-0-width-row-character}
\includegraphics[width=1.0\linewidth]{energy-vs-alpha-beta-0-width-row-character}
\caption{
Splitting and mixing in $t_{2g}$ manifold induced by octahedral tilting.
(a-c) 
$E_D, E_{\pm}$ and $E_0^{t_{2g}}$
as a function of $\alpha$
at 
$\beta=0$;
other parameters are the same as in Fig.~\ref{fig:t2g-alpha0}.
(d-f) Weight $W_{xy}$ of $xy$ orbital in 
$\ket{E_{\pm}}$ states 
shown in (a-c).
The $xy$ orbital predominantly
contributes to 
$\ket{E_{+}}$
at $q_A/q_o=1/2,1$ (a,b,d,e), 
but to   
$\ket{E_{-}}$
at $q_A/q_o=2$ (c,f).
}
\label{fig:t2g-beta0}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-vs-beta-alpha-10-width-row-beta-0-15}
\includegraphics[width=1.0\linewidth]{energy-vs-beta-alpha-10-width-row-beta-0-15}
\caption{
Splitting in $t_{2g}$ manifold due to octahedral rotation and tilting.
$t_{2g}$ energies as a function of $\beta$ at $\alpha=10^\circ$
at the same values of other parameters as in Figs.~\ref{fig:t2g-alpha0},\ref{fig:t2g-beta0}.
There are two anticrossings between the three states around $\beta\sim\alpha$.
The character of the states depends on 
the competition between the two angles and 
the two ion cages.
}
\label{fig:t2g-ge-Es}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/wts-energy-vs-beta-alpha-10-width-row-beta-0-15}
\includegraphics[width=1.0\linewidth]{wts-energy-vs-beta-alpha-10-width-row-beta-0-15}
\caption{
The dependence of the character of the $t_{2g}$ states
on the competitions between
$\alpha,\beta$ and 
A,B ion cages.
Weights of the three $t_{2g}$ orbitals in the three 
$t_{2g}$ eigenstates 
shown in
Fig.~\ref{fig:t2g-ge-Es}.
%(at the same parameters).
At $q_A=q_o/2$, B cage dominates and
$\ket{E_1}$ is predominantly $yz,zx$ at $\beta\sim 0\ll\alpha=10^\circ$, but 
becomes $\sim xy$ at $\beta\gtrsim10^\circ$ (a).
The situation flips when A cage dominates at $q_A=2q_o$ (d).
Similarly, the character of $\ket{E_3}$ at 
$q_A=q_o/2$
resembles 
that of $\ket{E_1}$ at $q_A=2q_o$  [compare (c) with (d)], and vice versa [compare (f) with (a)].
The character of the middle state $\ket{E_2}$ (b,e)
becomes $\sim xy$ only around the ``resonance'' region
$\beta\sim \alpha$.
}
\label{fig:t2g-ge-wts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/fig-data/energy-wts-eg}
\includegraphics[width=1.0\linewidth]{energy-wts-eg}
\caption{
Energies and characters of $e_g$ states.
$E_{\pm}$ and $E_0^{e_g}$
as a function of $\beta$ at $\alpha=10^\circ$, 
and the same parameters as in Fig.~\ref{fig:t2g-ge-Es}.
At $q_A=q_o/2$ (a,c) where B cage dominates,
$\alpha$ tends to make $\ket{E_{+}}$
more $x^2-y^2$ (see small $\beta$ region),
while $\beta$ tends to do the opposite 
(see large $\beta$ region).
The roles of $\alpha,\beta$ switch
at $q_A=2q_o$ (b,d), where A cage dominates.
We see that the splitting is relatively small as compared to $t_{2g}$ manifold.
Approximate results obtained by a second order expansion of the Hamiltonian given in Eqs.\ref{eq:eg-hiix}-\ref{eq:eg-E0}
are also shown for comparison.
}
\label{fig:e-wts-eg}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\linewidth]{/Users/maz/repos/perovskite/calc/analytical-structure-param/full-lattice-potential-tests/full-lattice-vs-AB}
\includegraphics[width=1.0\linewidth]{full-lattice-vs-AB}
\caption{
The effect of the potential of full ionic lattice.
$t_{2g}$ spectrum computed using the potential of full lattice 
and that obtained from only nearest AB cages (and the anion octahedron itself)
are
shown as a function of $q_A/q_o$ at 
$\alpha=0,\beta=20^\circ$, $\Delta_o=4q_o/15a$.
Full lattice only shifts the balance between the two cation cages towards A 
(i.e, the crossing point, where the two cages balance each other, moves to smaller $q_A$)
but keeps the qualitative picture the same.
}
\label{fig:t2g-lattice}
\end{center}
\end{figure}

@@

Of course, beside the crystal field we consider in our simplistic model,
in real materials,
 there would be other local effects like
electronic correlations (Hubbard and Hund's interaction)
and spin-orbit interaction, 
as well as band structure effects arising from
the electron hopping between the B ions.
The interplay between these and the crystal field (or JT) effect 
we discussed in this manuscript is also an interesting future work.
However, a few remarks are in order. 
Unlike the electronic correlations and spin-orbit interaction,
 the hybridization between 
the orbitals localized on different B sites are not
rotationally invariant.
Assuming the crystal field splitting to be the largest energy scale,
the orbitals in the $e_g$ and $t_{2g}$ manifolds of B sites are aligned with their respective octahedra,
and
the hybridization between the orbitals localized on different B sites
(and the electronic band structure arising from them)
are modified as the octahedra rotate.

To shed some light on the 
role of these band structure effects,
we consider the nearest neighbor hopping between the B sites
using the Slater-Koster method~\cite{SlaterKoster}
and
calculate the change in the energy of two configurations
$t_{2g}^1$ and $e_g^1$
as the octahedra rotate
in three cases: with either the crystal field or the hopping only and with both of these simultaneously.
Figure~\ref{fig:hybrid}(a)
shows the change in the energy 
as a function of $\beta$ for the first two cases
for the crystal-field-only (black curves, left y-axis;
$\alpha=0$, $q=2q_o$, $\Delta_o=4q_o/15a$)
and for the hopping-only case (red/gray curves, right y-axis).
We use Slater-Koster hopping parameters
$(t_{\sigma},t_{\pi},t_{\delta})=(0.15,-0.1,0.025)$~eV, 
$a=7$~bohr, $q_o=2$~e,
and consider only the crystal field of O octahedra that creates
$t_{2g}$ and $e_g$ manifolds.
While both $t_{2g}^1$ and $e_g^1$ configurations 
show a lowering of energy 
due to
the crystal field 
as $\beta$ 
increases---consistent with the results presented 
before---the energy of $e_g^1$ increases when only hopping is considered 
(the effect scales with the hopping parameters or the bandwidth).
This is due to a reduced hybridization leading to narrower $e_g$ bands.
In contrast, the energy of $t_{2g}^1$ decreases due band widening arising from 
an enhanced hybridization.
The crystal field and these band structure effects are not additive however.
The rotation induced splitting of the $t_{2g}$/$e_g$ manifolds
suppresses the effectiveness of the hopping in creating 
the dispersion in the energy states, 
and thus reduces the bandwidths and
affects the associated rotation induced energetic changes.

To show the validity regime of our crystal field model 
presented in this manuscript,
Fig.~\ref{fig:hybrid}(b)
presents  
the ratio of the
 change in the energy 
at $\beta=20^\circ$ (i.e., the energy difference from the cubic case)
at finite hopping $E(W)$ to that of our simplistic crystal field model $E(0)$
as a function of the average bandwidth $W$ of $t_{2g}$ and $e_g$ bands.
Here we scale the Slater-Koster parameters used for Fig.~\ref{fig:hybrid}(a)
to tune the bandwidth $W$.
We see that $E(W)/E(0)$ decreases from unity at $W=0$
to half its value for $t_{2g}^1$ configuration at $W/\Delta_o\gtrsim0.3$,
and even becomes negative for $e_g^1$ configuration around 
$W/\Delta_o\gtrsim0.15$ (showing the dominance of the band structure effects).
From this analysis, we conclude that the results of our crystal field model 
hold unambiguously only for narrower bands, $W/\Delta_o\ll1$.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.0\linewidth]{hybrid}
\caption{The effect of dispersion due to the electron hopping between B sites.
(a) Energy of two configurations $t_{2g}^1$ and $e_g^1$
as a function of $\beta$ when either the crystal field or the hopping is considered alone.
In the latter case, the energy changes are scaled with the bandwidth $W$.
(b) Considering both the crystal field and the hopping simultaneously,
the ratio of the change in energy with rotation $\beta$ at a finite hopping to that in the absence of hopping $E(W)/E(0)$
as a function of the average bandwidth $W$ 
for the same two configurations
$t_{2g}^1$ and $e_g^1$.
The results of the crystal field model presented in this manuscript 
hold in the small bandwidth limit where $E(W)/E(0)\sim 1$.
}
\label{fig:hybrid}
\end{center}
\end{figure}
